---
title: "Rethinking Walkthrough"
author: "Brian Holt"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE,echo=F}
knitr::opts_chunk$set(echo = TRUE)
library(rethinking)
library(tidyverse)
library(lubridate)
library(splines)
library(gridExtra)
```

# Chapter 4 Splines @splines

```{r}
data(cherry_blossoms)
d = cherry_blossoms
precis(d)
```

```{r}
with(d,plot(x=year,y=temp, col = rgb(red = 0, green = .5, blue = 01, alpha = 0.2),
cex=.5,ylab='March Temperature'))


```

```{r}
quantile(d$year)
range(d$year)
```

```{r}
d2 <- d[complete.cases(d$temp),]
num_knots <-5
knot_list <- quantile(d2$year,probs=seq(0,1, length.out=num_knots))
print(knot_list)
```

```{r}

B <- bs(d2$year,
        knots=knot_list[-c(1,num_knots)],
        degree = 10, intercept = T)
m4.7 <- quap(
  alist(
    D~dnorm(mu,sigma),
    mu <- a + B %*% w,
    a ~dnorm(100,10),
    w~ dnorm(0,10),
    sigma ~ dexp(1)
  ), data=list(D = d2$temp, B=B),
  start = list( w=rep(0, ncol(B)))
  )
```

```{r}
plot (NULL, xlim=range(d2$year),ylim=c(0,1),xlab='year',ylab='basis')
for(i in 1:ncol(B)){lines(d2$year,B[,i])}
```

```{r}
mu <- link(m4.7)
mu_PI <- apply(mu,2,PI,.97)
plot(d2$year, d2$temp,col=col.alpha(rangi2,.3),pch=16)
shade(mu_PI,d2$year,col=col.alpha("black",.5))
```

```{r}
post <- extract.samples(m4.7,n =20)
```

```{r}
plot(x=d2$year,y=d2$temp, col = rgb(red = 0, green = .5, blue = 01, alpha = 0.4),
cex=.5,ylab='March Temperature',ylim=(c(0,100)),xlim=(c(0,2000)))
 for (i in 1:20){
   curve(post$a[i]+post$sigma[i] * (x-mean(d2$temp,na.rm=T)),
         col=col.alpha('black',.3),add=T)
 }
```

# Waffle house

### DAG
#### fork or confounder
I learned this as the 3rd variable causing 2 others

richard uses rbern which might in his package, but rbinom works too.

Basically, he simulates how a 'z' variable influences 2 others, x and y

```{r,simulate binomial influence on other binom variables}

n <- 1000

z <- rbinom (n,1,prob=.5)
x <- rbinom(n,1,(1-z)*.1 + z*.9)
y <- rbinom(n,1,(1-z)*.1 + z*.9)

table(x,y)
cor(x,y)

cor(x[z==0],y[z==0])
cor(x[z==1],y[z==1])


```

```{r, sim a fork impacting gaussian variables}

# use z from above

cols <- c(4,2)
x.n <- rnorm(n,2*z-1) # 2*z-1 first makes 0's and 2's, then sub 1 to make a dummy code of -1 or 1
y.n <- rnorm(n,2*z-1)


plot(y.n~x.n,col=cols[z+1]) # the z+1 creates a vector, length of z, of 1's and 2's, not 0's and 1's.  The 1's correspond to the color code '4', while the 2's correspond to color code '2' because cols has 2 elements, place 1=4, place 2=2

abline(lm(y.n~x.n),lwd=3)
abline(lm(y.n[z==0]~x.n[z==0]),col=cols[1])
abline(lm(y.n[z==1]~x.n[z==1]),col=cols[2])

cor(x.n,y.n)

cor(x[z==0],y[z==0])
cor(x[z==1],y[z==1])

```

```{r, the pipe}

#x -> z -> y, x influences y are transmitted thu z

#x goes to z, contanimaints z in effecting y#


n <- 1000

x <- rbinom (n,1,prob=.5)
z <- rbinom(n,1,(1-x)*.1 + x*.9)
y <- rbinom(n,1,(1-z)*.1 + z*.9)

table(x,y)
cor(x,y)

cor(x[z==0],y[z==0])
cor(x[z==1],y[z==1])


#...it's the same as the fork just..the causal steps are different
```

```{r, rnorm with a vector of probs}
n <- 100
age <- rnorm(n)

```


```{r}
data(WaffleDivorce)
w <- WaffleDivorce
str(w)
precis(w)
```

```{r}

cor(w$MedianAgeMarriage,w$Divorce)
fit.1<- lm(Divorce~Marriage, w)
fit.2<- lm(Divorce~MedianAgeMarriage, w)
fit.3 <- lm(Divorce~MedianAgeMarriage+Marriage, w)
summary(fit.3)
par(mfrow=c(2,2))
plot(w$Divorce~w$Marriage)
abline(fit.1)
plot(w$Divorce~w$MedianAgeMarriage)
abline(fit.2)
plot(w$MedianAgeMarriage,w$Marriage)
par(mfrow=c(1,1))
```

reminders: \# fit.1\<- lm(Divorce\~Marriage, w) \# fit.2\<- lm(Divorce\~MedianAgeMarriage, w) \# fit.3 \<- lm(Divorce\~MedianAgeMarriage+Marriage, w) \# fit.4 \<- lm(Marriage\~MedianAgeMarriage, w)

```{r}
fit.4 <- lm(Marriage~MedianAgeMarriage, w)
summary(fit.4)
fit.4$residuals
```

```{r}
w %>%  ggplot(aes(x=MedianAgeMarriage,y=Marriage))+
  geom_point() +geom_smooth(method='lm') -> plot1

w %>% mutate(Mar.MedAge.resid= fit.4$residuals) %>%  ggplot(aes(x=Mar.MedAge.resid,y=Divorce))+
  geom_point() + geom_smooth(method='lm')-> plot2
```

```{r}
#Wondering about interactions

fit.5 <- lm(Divorce~MedianAgeMarriage*Marriage, w)
summary(fit.5)
plot(y=w$Divorce,fit.5$residuals)

```

```{r}
grid.arrange(plot1,plot2,ncol=2)

```

```{r}
n_samples <- 1000
p <- rep( NA , n_samples )
p[1] <- 0.5
W <- 6
L <- 3
for ( i in 2:n_samples ) {
p_new <- rnorm( 1 , p[i-1] , 0.1 )
if ( p_new < 0 ) p_new <- abs( p_new )
if ( p_new > 1 ) p_new <- 2 - p_new
q0 <- dbinom( W , W+L , p[i-1] )
q1 <- dbinom( W , W+L , p_new )
print(q0)
print(q1)
cat ("\nhere is the ratio\n")
cat(q1/q0,"\n\n")
cat("here is the p-new.......",p_new,'\n')
ran.unif<- runif(1)
cat("here is the runif(1).....",ran.unif,"\n")
p[i] <- ifelse( ran.unif < q1/q0 , p_new , p[i-1] )
}
```

```{r}
dens( p , xlim=c(0,1) )
curve( dbeta( x , W+1 , L+1 ) , lty=2 , add=TRUE )
```


